\textbf{Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-07-15 submission/manuscript.tex
Sent to Pat

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Evaluation of machine learning methods for 16S rRNA gene data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Evaluation of machine learning methods for 16S rRNA gene data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

Begüm D. Topçuoğlu\({^1}\), Nick Lesniak\({^1}\), Jenna Wiens\({^2}\),
Mack Ruffin\({^3}\), Patrick D. Schloss\textsuperscript{1\(\dagger\)}

@@

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex,
high-dimensional and uneven within and between individuals with the same
disease. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for differences in health outcomes.

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. They learn from
existing data to predict the outcomes of new data and allow us to infer
on the reasons underlying that prediction. Therefore, researchers have
started to explore the utility of ML models that use microbiota
associated biomarkers to predict human health and to understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
(1--11). However, the field's use of ML lacks clarity and consistency in
which methods are used and how these methods are implemented (12, 13).
More notably, we commonly see flawed ML practices such as (1) using ML
pipelines where there is no seperate held-out test dataset to evaluate
model performance or (2) reporting few or only the best outcomes of
cross-validation. Even when there are seperate testing sets to evaluate
model performance, there are large differences between cross-validation
and testing performances that indicate overfitting as well as large
confidence intervals for testing performances (4, 14--20).

Moreover, there is a lack of discussion on why a particular ML model is
utilized. Recently, there is a trend towards using more complex ML
models such as random forest, extreme gradient boosting and neural
networks without a discussion on if and how much model interpretibility
is necessary for the study (11, 21--23). Black box machine learning
models are not inherently interpretable and require posthoc explanations
to determine the feature importances in making a prediction. These
explanations can be misleading and at times unreliable when making
high-stake decisions about patient health (24). The models we develop
for healthcare, both to predict disease and to understand underlying
reasons behind that prediction, should be transparent and accountable
(25).

The lack of transparency on model selection and interpretation as well
as flawed modeling methods negatively impact model validity and
reproducibiliy. We need to strive toward better machine learning
practices by (1) implementing rigourous machine learning pipelines and
(2) selecting ML models that reflect the goal of the study as it will
inform our expectations of model accuracy, complexity, interpretibility
and computational efficiency.To showcase a rigorous ML pipeline and to
shed light on how ML model selection can affect modeling results, we
performed an empirical analysis comparing several different ML models
using the same dataset and the same ML pipeline. We used a previously
published colorectal cancer (CRC) study (3) which had fecal 16S rRNA
gene sequences of 490 patients. The human-associated human microbiome is
hypothesized to directly contribute to the development of CRC and fecal
16S rRNA gene sequences have been used to detect it (1, 3, 4, 26). We
built seven ML models using fecal 16S rRNA gene sequences to predict
healthy patients versus patients with colorectal lesions that were
identified by colonoscopy as screen relevant neoplasias (SRN). The study
had 261 normal and 229 SRN samples. We established modeling pipelines
for L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels, a decision
tree, random forest and XGBoost. We used the area under the receiver
operating characteristic curve (AUROC) as a metric for determining the
predictibve performance of each model. The AUROC ranges from 1.0, where
the model perfectly distinguishes between cases and controls, to 0.50,
where the model's predictions are no different from random chance. The
median test AUROC varied from 0.601 to 0.695. Random forest had the
highest median AUROC for detecting SRN. Despite its simplicity, the
L2-regularized logistic regression was second best in predictive
performance. In terms of computational efficiency, L1 SVM with linear
kernel trained the fastest (0.202 hours, std ± 0.028), while XGBoost
took the longest (155.104 hours, std ± 0.959). We also found that
depending on how the data is split to create a held-out test set, the
AUROC values of a ML model can vary up to 0.295 which highlighted the
importance of performing many randomized modeling runs. This study
established standards for microbiome-associated ML models and
underscored the importance of model selection.

@@

We used a cohort of 490 patients with 261 cases of SRN. For each
patient, we had 6920 features (fecal bacterial abundances) and a
two-class label that defines their colorectal health (having colorectal
lesions that are identified as SRN or normal). All the cases were
independently labeled through colonoscopies. We established modeling
pipelines for a binary prediction task with L2-regularized logistic
regression, L1 and L2 support vector machines (SVM) with linear and
radial basis function kernels, a decision tree, random forest and
extreme gradient boosted decision tree (XGBoost) to emphasize the
differences in model accuracy, complexity, interpretibility and
computational efficiency due to model selection.

For regularized logistic regression and SVM with linear kernel we used
L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data lead to a sparser solution
(i.e., force many coefficients to zero). Finally, to explore the
potential for non-linear relationships among features and the outcome of
interest, we trained tree based models, decision tree, random forest and
XGboost, as well as an SVM with non-linear kernel.

We established a ML pipeline where we train and validate each of the
seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection (i.e.~model selection), and the test set was
used for evaluation purposes. Validation of hyperparameter selection was
performed using repeated five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over 100 randomizations and selected the best performing
hyperparameter setting in terms of AUROC metric and trained the full
training/validation dataset {[}Figures S1 and S2{]}. We then used the
held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model performance {[}Figure 1{]}.

\textbf{Discriminative performance and generalizability of the seven
models.}

We evaluated the predictive performances of seven binary classification
models when applied to held-out test data using AUROC metric {[}Figure
2{]}. Random forest had significantly higher test AUROC values than the
other models for detecting SRNs when AUROC values were compared to the
other six by Wilcoxon rank sum test (p \textless{} 0.01). The median
AUROC of the random forest model was 0.695 (IQR 0.044). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel AUROC values were not significantly different from
one another. They had median AUROC values of 0.68 (IQR 0.055), 0.679
(IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1
SVM with linear kernel and decision tree had significantly lower AUROC
values than the other ML models with median AUROC of 0.65 (IQR 0.066)
and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.

For each model, we compared the median cross-validation AUROC to the
median testing AUROC. The difference between the two should be low to
suggest the model is not overfitting despite the large number of
features. The largest difference between the two was 0.021 in L1 SVM
with linear kernel, followed by SVM with radial basis function kernel
and decision tree with a difference of 0.007 and 0.006, respectively
{[}Figure 2{]}.

We reported the testing AUROC values over 100 randomizations of the
initial data-split. The results showed that depending on the data-split,
the testing AUROC values showed great variability {[}Figure 2{]}. The
testing AUROC values within each model varied 0.23 on average across the
seven models. For instance, the lowest AUROC value of the random forest
model was 0.59 whereas the highest was 0.81.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel and L2 logistic regression using the absolute
feature weights of the trained models. We ranked the absolute weights of
all the OTUs for each data-split {[}Figure 3{]}. We calculated the
median ranks of these features over the 100 data-splits. In the three
linear models, OTUs that had the largest median ranks and drove the
detection of SRNs belonged to families \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
OTU00015, OTU00768, OTU00822, OTU00609), genera \emph{Gamella}
(OTU00426) and genera \emph{Peptostreptococcus} (OTU00367) {[}Figure
3{]}. Some of the OTUs with the highest ranks were shared among the
linear models.

We explained the feature importances in non-linear models; SVM with
radial basis kernel, decision tree, random forest and XGBoost, using a
method called permutation importance on the held-out test data.
Permutation importance analysis is where we randomly permute
non-correlated features individually and groups of highly correlated
features together. We then calculate how much the predictive performance
of the model (i.e AUROC values) decrease when each OTU or group of OTUs
is permuted randomly. We ranked the OTUs based on how much they
decreased the median testing AUROC; the OTU with the largest decrease
ranking highest. The top 5 OTUs with the largest negative impact on
testing AUROC overlapped in tree-based models {[}Figure 4{]}.
Specifically, permuting \emph{Peptostreptococcus} (OTU00367) abundances
randomly, dropped the predictive performances the most in all tree-based
methods {[}Figure 4{]}. Decision tree, random forest and XGBoost models'
predictive performance dropped from 0.6 median AUROC to 0.52, from 0.69
to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.

To highlight the differences between the two interpretation methods, we
used permutation importance to interpret linear models as well {[}Figure
S3{]}. L1-regularized SVM with linear kernel picked out some of the same
OTUs (OTU00822, OTU01239, OTU00609) as important in the two
interpretation methods; feature rankings based on weights {[}Figure 3{]}
and permutation importance {[}Figure S3{]}. Similarly, L2-regularized
SVM and L2-regularized logistic regression picked out some of the same
OTUs in both interpretation methods, OTU00659 and OTU00012,
respectively. However, for all the linear models, the rankings of these
features were different due to the collinearity in microbial
communities.

@@

As the complexity of a ML model {[}Table S1{]} and the number of tuned
hyperparameter settings increased {[}Figures S1-S2{]}, its training
times increased as well {[}Figure 5{]}. Linear models trained faster
than non-linear models. L1 and L2 SVM with linear kernel and L2 logistic
regression had training times of 0.2 hours, (std ± 0.03), 0.2 hours,
(std ± 0.02), and 0.2 hours, (std ± 0.02), respectively. Whereas, a
decision tree, SVM with radial basis function kernel, random forest and
xgboost had training times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ±
8.8), 83.2 hours, (std ± 11.3) and 155.1 hours, (std ± 1), respectively
{[}Figure 5{]}.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We set-up standards
for developing and evaluating ML models for microbiome data. First, we
used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the difference between
cross-validation and test performance is low, this suggets the models
are not overfit and that they will perform similar with similar data. In
all seven models, the difference between median cross-validation and
testing AUROC values did not exceed 0.021 which suggests that these
models are generalizable and can be used to test similar new data.
Second, we performed the initial 80\%-20\% random datasplit 100 times in
our ML pipeline. The randomization of the initial data-split to create a
held-out test set is a crucial step in the ML pipeline to develop robust
ML models and to report reliable performance metrics for a ML model.
Depending on how the data is split, there is the chance of being
overoptimistic about the predictive performance of a model. In our
study, we showed that there was variability in AUROC values between
different random data-splits in each of the models we tested. Our
results showed that the testing AUROC values varied 0.23 on average
between different data-splits. Third, we used the AUROC metric instead
of accuracy in our study to evaluate the predictive performance of the
ML models. AUROC is always random at the value 0.5 and is a robust
metric when a dataset is imbalanced. We also performed a full grid
search for hyperparameter settings when building a ML model. Default
hyperparameter settings in previously developed ML packages in R,
Python, and Matlab programming languages are inadequate for effective
application of classification algorithms and need to be optimized for
each new dataset used to generate a model. In the example of
L1-regularized SVM with linear kernel {[}Figure S1{]}, the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

In this study, we benchmarked seven ML models with different
classification algorithms to show that we should use ML models based on
the goal of the study and our expectations of predictive performance,
interpretibility and computational burden. Microbiome studies use ML
models with a classification task to learn the training data to assign
labels, such as healthy or not, to new data, but also to learn which
features are important to discriminate between labels. If the goal of a
study is not just to predict a health outcome but also to learn the
ecology behind a disease and to identify microbial biomarkers of a
disease, then the ML model has to be interpretable. In terms of
predictive performance, random forest model had testing AUROC values
statistically significantly larger than others. However the second best
model was L2-regularized logistic regression following random forest
with a median AUROC difference of only 0.015. In terms of
interpretability, random forest was a complex ML model and it could only
be explained using methods such as permutation importance. On the other
hand, L2-regularized logistic regression was easier to interpret
(i.e.~ranking absolute regression coefficients of the trained model).

Even with interpretable models such as L2-regularized logistic
regression, we need to be careful with how we treat the information we
get from the models. In this study we used two different methods to
interpret our linear models; ranking each OTU by (1) their absolute
weights in the trained models and (2) their impact on the predictive
performance based on permutation importance. We observed differences in
the OTU rankings between the two interpretation methods due to
collinearity in the dataset. Collinearity in a microbial dataset is when
one OTU is highly correlated with another OTU which causes feature
weights to not be unique. The feature weights of correlated OTUs are
influenced by one another which makes it difficult to interpret the ML
model. To avoid misinterpreting the models we can; (1) use ML models
that has highly correlated OTUs removed, (2) use the highly ranked OTUs
that are correlated with other OTUs to generate hypotheses about the
ecology of the disease and test them with followup experiments.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Best practices for applying machine learning to bacterial 16S
rRNA gene sequencing data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Jenna
Wiens\({^2}\), Mack Ruffin\({^3}\), Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

@@

Advances in sequencing technology and decreases in the costs of
generating 16S rRNA gene sequences have allowed rapid exploration of the
human microbiome and its health implications. Currently, the human
microbiome field is growing at an unprecedented rate and there is an
increasing demand for methods that identify associations between
microbiome members and human health. However, this is difficult as human
associated microbial communities are remarkably complex,
high-dimensional and uneven within and between individuals with the same
disease. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for differences in health outcomes.

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. They learn from
existing data to predict the outcomes of new data and allow us to infer
on the reasons underlying that prediction. Therefore, researchers have
started to explore the utility of ML models that use microbiota
associated biomarkers to predict human health and to understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
The high-stake task of predicting the correct diagnosis with high
confidence relies on a ML model that is built with rigorous methods.
However, the field's use of ML lacks clarity and consistency in which
methods are used and how these methods are implemented (12, 13). More
notably, we commonly see flawed ML practices such as using ML pipelines
where there is no seperate held-out test dataset to evaluate predictive
performance or reporting few or only the best outcomes of
cross-validation. Even when there are seperate testing sets to evaluate
predictive performance, there are large differences between
cross-validation and testing performances and large confidence intervals
in testing performances (4, 14--20). These practices prevent the
development of generalizable models, where the model makes accurate
predictions with newly acquired data as well as it does with the
training data.

Moreover, there is a lack of discussion on why a particular ML model is
utilized and a lack of emphasis on the strengths and weaknesses of it.
Recently, there is a trend towards using more complex ML models such as
random forest, extreme gradient boosting (XGBoost) and neural networks
without a discussion on if and how much model interpretibility is
necessary for the study (11, 21--23). These models are also called black
box ML models because they are not inherently interpretable and require
posthoc explanations to determine the feature importances in making a
prediction. These explanations can be misleading and at times unreliable
when making high-stake decisions about patient health (24). The models
we develop for healthcare, both to predict disease and to understand
underlying reasons behind that prediction, should be transparent and
accountable (25).

The lack of transparency on model selection and interpretation as well
as flawed modeling methods negatively impact model validity and
reproducibility. We need to strive toward better machine learning
practices by (1) implementing rigorous machine learning pipelines and
(2) selecting ML models that reflect the goal of the study as it will
inform our expectations of predictive performance, complexity,
interpretability and computational efficiency. To showcase a rigorous ML
pipeline and to shed light on how ML model selection can affect modeling
results, we performed an empirical analysis comparing several different
ML models using the same dataset and the same ML pipeline. We used a
previously published colorectal cancer (CRC) study (3) which had fecal
16S rRNA gene sequences of CRC patients. The human microbiome is
hypothesized to directly contribute to the development of CRC and fecal
16S rRNA gene sequences have been used to detect it (1, 3, 4, 26). We
built seven ML models using fecal 16S rRNA gene sequences to predict
healthy patients versus patients with colorectal lesions that were
identified by colonoscopy as screen relevant neoplasias (SRN). We
established modeling pipelines for three linear models with different
forms of regularization; L2-regularized logistic regression, L1 and
L2-regularized support vector machines (SVM) with linear kernel. We also
developed four non-linear models; SVM with radial basis function kernel,
a decision tree, random forest and XGBoost. We compared the predictive
performance, interpretability and computational efficiency of the seven
models to highlight the importance of model selection. We established
standards for ML pipeline construction, predictive performance
evaluation and model interpretation for microbiome-associated ML models
that have a binary prediction task.

@@

We first determined the dataset and the ML models to use in our study.
We used a previously published study on a CRC cohort of 490 patients
with 261 cases of SRN. For each patient, we had 6920 features (fecal
bacterial abundances) and a two-class label that defines their
colorectal health (normal or SRN colorectal lesions as defined by
colonoscopies). We established modeling pipelines for a binary
prediction task with L2-regularized logistic regression, L1 and L2
support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and extreme gradient boosted
decision tree (XGBoost).

We used ML models with different classification algorithms and
regularization parameters. For regularized logistic regression and SVM
with linear kernel, we used L2 regularization to keep all potentially
important features. For comparison, we also trained an L1 regularized
SVM model with linear kernel. L1-regularization on microbiome data lead
to a sparser solution (i.e., force many coefficients to zero). Finally,
to explore the potential for non-linear relationships among features and
the outcome of interest, we trained tree based models, decision tree,
random forest and XGBoost, as well as an SVM with non-linear kernel.

We established a ML pipeline where we train and validate each of the
seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection, and the test set was used for evaluation
purposes.

Hyperparameters are the rules that are learned from the training data in
a classification algorithm. When they are tuned over a full grid search
and selected by validation, they make the ML model better are
predicting. We selected hyperparameter settings using repeated five-fold
cross-validation on the training/validation set {[}Figure 1{]}. We chose
the best hyperparameter setting for each model based on its predictive
performance on the validation set using the area under the receiver
operating characteristic curve (AUROC) metric {[}Figure S1 and S2{]}.
The AUROC ranges from 1.0, where the model perfectly distinguishes
between cases and controls, to 0.50, where the model's predictions are
no different from random chance. Similar to the initial data-split,
five-fold cross-validation for hyperparameter selection was also
stratified to maintain the overall label distribution on the training
and validation sets. The cross-validation of each hyperparameter setting
was repeated over 100 randomizations to get a robust reading of
predictive performance.

We then trained the full training/validation dataset with the selected
hyperparameters. We used the held-out test set to evaluate the
predictive performance of each ML model. The data-split, hyperparameter
selection, training and testing steps were repeated 100 times to get a
reliable and robust reading of model performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.}

We evaluated the predictive performances of seven binary classification
models when applied to held-out test data using the AUROC metric
{[}Figure 2{]}. Random forest had significantly higher test AUROC values
than the other models for detecting SRNs when AUROC values were compared
to the other six by Wilcoxon rank sum test (p \textless{} 0.01). The
median AUROC of the random forest model was 0.695 (IQR 0.044).
L2-regularized logistic regression, XGBoost, L2-regularized SVM with
linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Random forest had the highest median AUROC for detecting
SRN. Despite its simplicity, the L2-regularized logistic regression was
second best in predictive performance.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. The difference
between the two should be low to suggest the model is not overfitting
despite the large number of features. The largest difference between the
two was 0.021 in L1-regularized SVM with linear kernel, followed by SVM
with radial basis function kernel and decision tree with a difference of
0.007 and 0.006, respectively {[}Figure 2{]}. We also reported the
testing AUROC values over 100 randomizations of the initial data-split.
The testing AUROC values within each model varied 0.23 on average across
the seven models. For instance, the lowest AUROC value of the random
forest model was 0.59 whereas the highest was 0.81. These results showed
that depending on the data-split, the testing AUROC values showed great
variability {[}Figure 2{]}.

@@

Interpretability is the degree to which humans can understand the
reasons behind a model prediction (27). Because we often use ML models
not just to predict a health outcome but also to learn the ecology
behind a disease, model interpretation becomes crucial for microbiome
studies. The ML models we built using L2-regularized logistic
regression, L1 and L2 support vector machines (SVM) with linear and
radial basis function kernels, a decision tree, random forest and
XGBoost decrease in interpretability as they increase in complexity. In
this study we highlighted two methods to interpret models with varying
complexity.

We interpreted linear models (L1 and L2-regularized SVM with linear
kernel and L2-regularized logistic regression) using the absolute
feature weights of the trained models. We ranked the absolute weights of
all the OTUs for each data-split {[}Figure 3{]}. We calculated the
median ranks of these features over the 100 data-splits. In the three
linear models, OTUs that had the largest median ranks and drove the
detection of SRNs belonged to families \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
OTU00015, OTU00768, OTU00822, OTU00609), genera \emph{Gamella}
(OTU00426) and genera \emph{Peptostreptococcus} (OTU00367) {[}Figure
3{]}. Some of the OTUs with the highest ranks were shared among the
linear models.

We explained the feature importances in non-linear models; SVM with
radial basis kernel, decision tree, random forest and XGBoost, using a
method called permutation importance on the held-out test data.
Permutation importance analysis is a posthoc explanation of the model
where we randomly permute non-correlated features individually and
groups of highly correlated features together. We then calculate how
much the predictive performance of the model (i.e AUROC values) decrease
when each OTU or group of OTUs is permuted randomly. We ranked the OTUs
based on how much they decreased the median testing AUROC; the OTU with
the largest decrease ranking highest. The top 5 OTUs with the largest
negative impact on testing AUROC overlapped in tree-based models
{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
(OTU00367) abundances randomly, dropped the predictive performances the
most in all tree-based methods {[}Figure 4{]}. Decision tree, random
forest and XGBoost models' predictive performance dropped from 0.6
median AUROC to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}.

To highlight the differences between the two interpretation methods, we
used permutation importance to interpret linear models as well {[}Figure
S3{]}. L1-regularized SVM with linear kernel picked out some of the same
OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings
based on weights {[}Figure 3{]} and permutation importance {[}Figure
S3{]}. Similarly, L2-regularized SVM and L2-regularized logistic
regression picked out some of the same OTUs in both interpretation
methods, OTU00659 and OTU00012, respectively. However, for all the
linear models, the rankings of these features were different due to the
collinearity in microbial communities.

@@

We compared the training times of the seven ML models. As the complexity
of a ML model {[}Table S1{]} and the number of tuned hyperparameter
settings increased {[}Figures S1-S2{]}, its training times increased as
well {[}Figure 5{]}. Linear models trained faster than non-linear
models. L1 and L2 SVM with linear kernel and L2-regularized logistic
regression had the shortest training times with 0.2 hours, (std ± 0.03),
0.2 hours, (std ± 0.02), and 0.2 hours, (std ± 0.02), respectively.
Whereas, a decision tree, SVM with radial basis function kernel, random
forest and XGBoost had training times of 4.4 hours, (std ± 0.3), 59.6
hours, (std ± 8.8), 83.2 hours, (std ± 11.3) and 155.1 hours, (std ± 1),
respectively {[}Figure 5{]}.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We built on others'
work in the microbiome field to set-up standards for developing and
evaluating ML models for microbiome data (28--30). First, we used a
held-out test set to illustrate the difference between cross-validation
and testing AUROC values. When the difference between cross-validation
and test performance is low, this suggest the models are not overfit and
that they will perform similar with similar data. In all seven models,
the difference between median cross-validation and testing AUROC values
did not exceed 0.021 which suggests that these models are generalizable
and can be used to test similar new data. Second, we performed the
initial 80\%-20\% random datasplit 100 times in our ML pipeline. The
randomization of the initial data-split to create a held-out test set is
a crucial step in the ML pipeline to develop robust ML models and to
report reliable performance metrics. Depending on how the data is split,
there is the chance of being overoptimistic about the predictive
performance of a model. In our study, we showed that there was
variability in AUROC values between different random data-splits in each
of the models we tested. Our results showed that the testing AUROC
values varied 0.23 on average between different data-splits. Third, we
used the AUROC metric in our study to evaluate the predictive
performance of the ML models. AUROC is always random at the value 0.5
and is a robust metric when a dataset is imbalanced. We also performed a
full grid search for hyperparameter settings when building a ML model.
Default hyperparameter settings in previously developed ML packages in
R, Python, and Matlab programming languages are inadequate for effective
application of classification algorithms and need to be optimized for
each new dataset used to generate a model. In the example of
L1-regularized SVM with linear kernel {[}Figure S1{]}, the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

In this study, we benchmarked seven ML models with different
classification algorithms to show that we should use ML models based on
the goal of the study and our expectations of predictive performance,
interpretability and computational burden. Microbiome studies use ML
models with a classification task to learn the training data to assign
labels, such as healthy or not to new data, but also to learn which
features are important to discriminate between labels (2--11). Our
results show that if the goal of a study is to learn the ecology behind
a disease and to identify microbial biomarkers, we could create ML
models that are inherently interpretable without losing predictive
power. In terms of predictive performance, random forest model had the
best testing AUROC values compared to the other 6 models. However, the
second best model was L2-regularized logistic regression with a median
AUROC difference of only 0.015 compared to random forest. In terms of
interpretability, random forest was a complex ML model and could only be
explained using methods such as permutation importance. On the other
hand, L2-regularized logistic regression was easier to interpret
(i.e.~ranking absolute regression coefficients of the trained model).

Even with interpretable models such as L2-regularized logistic
regression, we need to be careful with how we treat the information we
get from the models. In this study we used two different methods to
interpret our linear models; ranking each OTU by (1) their absolute
weights in the trained models and (2) their impact on the predictive
performance based on permutation importance. We observed differences in
the OTU rankings between the two interpretation methods due to
collinearity in the dataset. Collinearity in a microbial dataset is when
one OTU is highly correlated with another OTU which causes feature
weights to not be unique. The feature weights of correlated OTUs are
influenced by one another which makes it difficult to interpret the ML
model. To avoid misinterpreting the models we should use the highly
ranked correlated OTUs to generate hypotheses about the ecology of the
disease and test them with follow-up experiments.

There are other criteria when choosing ML models such as the
computational burden of developing it and the sample size. In terms of
computational burden, random forest model trained each data-split in
83.2 hours whereas L2-regularized logistic regression trained in 12
minutes. The generalization performance of ML models depends on sample
size. The more complex the model, the more data it will need. The
dataset we used for our study had 490 samples, however microbiome
studies that have smaller sample sizes would benefit from using less
complex models such as L2-regularized logistic regression.

This study highlights the need to make educated choices at every step of
developing a ML model with microbiome data. Model selection should be
done with a solid understanding of model complexity and
interpretability, rigorous ML pipelines should be built with
cross-validation for hyperparameter tuning and with a held-out test set
for evaluating predictive performance and models should be interpreted
while considering collinearity in datasets. The right methods will help
us achieve the level of validity and accountability we want from models
built for patient health.
